Calculating the Solvent Exposed Surface Area of Proteins with Intersecting Spheres

### 2020-06-06 Paper.tex
More Edits.

\documentclass{article}
\usepackage{amssymb, amsfonts, amsmath, geometry, verbatim, caption}
\geometry{letterpaper}
\usepackage[parfill]{parskip}    		
\usepackage{graphicx}
\graphicspath{{./}}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage[utf8]{inputenc}

@@

All points on $a_i$ within $d_i$ of the $P_{iq}$ are buried, and those bits in $m_{iq}$ are set to zero, with the rest set to 1. This string can be AND'ed with $b_i$ to produce a new $b_i$. This can be done for all spheres that intersect $a_i$ for each sphere in turn. The key insight that cuts down on the $nkl$ calculations is the fact that the $m_{iq}$ can be stored and retrieved later if the same string is ever needed. This is very likely to occur, because the $d_{iq}$ can be made independent of $r_i$ by removing the factor $r_i$. Then $d_{iq}=\sqrt{2(1-\cos\theta_{iq})}$. Now we can calculate $m_{iq}$ for the range of $d_{iq}$ from 0-2. Once we store these masks, we will have covered all possible overlaps between spheres because we have gone through all angles $\theta_{iq}$.
This provides a method of calculating the SASA of the entire protein. We can simply look at every intersection with a given atom and lookup the appropriate mask (based on the angle $\theta_{iq}$. Then we scale it by the radius of the atom we are looking at. Then we can perform our AND operations and find the SASA for a given atom. Iterating over all atoms gives us our answer.
\subsection{Richmond's Algorithm and the Gauss-Bonnet Theorem}
In a previous paper written by Richmond (1984), a different strategy is used to find the solvent exposed surface area of the atoms. He still represents the atoms as spheres with definite radius with centers at the atomic coordinates, but instead of using voxels, he made use of the Gauss-Bonnet theorem, a fundamental theorem in differential geometry which connects the topological properties of a manifold (specifically its Euler Characteristic) with its geometric properties (its curvature). A standard statement of the theorem is

@@

\begin{figure}[h!]
\caption{The dual complex ($K$) of the Delaunay Triangulation only includes those power cells that have a non-empty intersection.$\lbrack1\rbrack$}
\centerline{\includegraphics[width=3in]{Figure9}}
\end{figure}

@@

\begin{figure}[h!]
\caption{The star of $K$ and link of the star are shown.$\lbrack1\rbrack$}
\centerline{\includegraphics[width=3in]{Figure5}}
\end{figure}

@@

bd $F$ means the boundary of $F$. This equation shows that for the space-filling diagram of the two intersecting spheres, $\sigma_{ij}$ is the fraction of the exposed surface that's on the power cells of both of the spheres.
Using the dual complex, we can further specify the equation for $\sigma_{ij}$ in terms of two intersecting circles, which will become useful later:

@@

The general equation of a sphere ($x^2 + y^2 +z^2 = r^2$) can be used to construct equations for $S_i$ and $S_j$:
\begin{align*}
&x^2 + y^2 +(z-z_i)^2 = \zeta_i^2 \\
&x^2 + y^2 +(z-z_j)^2 = \zeta_j^2 \\
&\text{Now we can combine the equations:} \\
&\zeta_i^2-\zeta_j^2 =(z-z_i)^2-(z-z_j)^2=-2z_i +2z_j +z_i^2 - z_j^2\\
& \implies -2z_i +2z_j +z_i^2 - z_j^2 = (\zeta_i)^2-(\zeta_j)^2\\
\end{align*}
Using this logic, we can find the same equation holds true if we consider the set of balls for each sphere (assuming they have they same radii), as they have the same center as the spheres. So:
\begin{align*}
&-2z_i +2z_j +z_i^2 - z_j^2 = \varrho_i^2-\varrho_j^2 \\
&\text{From the above two expressions, we can state that:} \\
&\zeta_i^2-\zeta_j^2 = \varrho_i^2-\varrho_j^2\\
&\text{In the left term, we can factor out $\zeta_i^2-\zeta_j^2$ to get:} \\
&\zeta_i^2-\zeta_j^2 =(\zeta_i-\zeta_j)(\zeta_i+\zeta_j) = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{So we can state that:} \\
&\varrho_i^2 - \varrho_j^2  = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{Now we can find what $\zeta_i$ and $\zeta_j$ really are} \\
&\varrho_i^2 - \varrho_j^2)/\zeta_{ij} + \zeta_j = \zeta_i \\
&\text{We know that $\zeta_j = \zeta_{ij} - \zeta_i$, so just substitute this expression in:} \\
&\zeta_i = \frac{1}{2}\left(\zeta_{ij}+\frac{\varrho_i^2-\varrho_j^2}{\zeta_{ij}}\right) \\
&\text{A similar statement can be made for $\zeta_j$, following the steps we carried out above:} \\
&\zeta_j = \frac{1}{2}\left[\zeta_{ij}+\frac{\varrho_j^2-\varrho_i^2}{\zeta_{ij}}\right]
\end{align*}

@@

\subsection{Distance Preserving Motion}
In distance preserving motion, one sphere rotates about the other. Imagine three spheres interesting at two points, $p$ and $q$ (Figure 9). This time, $F$ is going to include all the points contained in all three spheres. It is assumed that if the spheres intersect at two points then one sphere contains the other two. In this case, it is $S_j$ that engulfs $S_i$ and $S_k$. The caps, or tops of the spheres, are considered. From a previous section, we know that $B_{ijk}$ is considered the line segment of three intersecting spheres. In this case, $B_{ijk}$ is the line segment containing $p$ and $q$. The caps $C_{ij} = B_i \cup S_j$ and $C_{kj} = B_k \cup S_j$ protrude from the surface of $S_j$. The parameters obtained from this diagram allows us to find the area similar to how we found it for the direction preserving motion. This model says that as you move the two caps apart, keeping their centers in the same position, you will gain an area equal to that of a spherical rectangle. Essentially, the caps are rotating in place. It might be easy to think about how diameter of the caps can represent the height of this “rectangle” while the distance between the caps signifies its width.
\begin{figure}[h!]
\caption{Three intersecting spheres at two points: $S_i$, $S_k$ and $S_j.\lbrack1\rbrack$}
\centerline{\includegraphics[width=3in]{Figure7}}
\end{figure}

@@

We also plotted the distribution of the SASA values for each calculator, and while the don't look too different in this example, more rigorous testing with a wider range of proteins has shown that \textsc{alphamol} gives far more accurate values than other algorithms, which is why it is incorporated into Rosetta today.
\begin{figure}
\caption{SASA Histograms for the Two Algorithms}
\centerline{\includegraphics[width=3in]{sasa}}
\end{figure}

\section{Conclusion}
It can be seen that SASA is a critical quantity to be able to compute for the purposes of structural biology. On a per-atom basis, it is one metric that can be used to determine if an atom is making a hydrogen bond, and by extension if the structure is energetically favorable. For protein-protein interfaces, large SASA values at the interface usually mean the proteins are more likely to bind because there is a larger binding surface. Metrics like SASA are useful because they don't require your structure to look similar to any existing PDB in order to determine if the structure is likely to be stable. This is especially important in the field of protein engineering, where there might not be any structures to compare your PDB to, and thus having independent metrics is critical to filtering out possible \textit{de novo} protein designs from sequences that won't take on any definite conformation. Thus much time has been spent on applying mathematical and computational ideas to rigorously defining and calculating SASA, in this paper we have explored a few of the most prominent methods.

>>>>

\documentclass{article}
\usepackage{amssymb, amsfonts, amsmath, verbatim, caption}
\usepackage{geometry}
\geometry{letterpaper}
\usepackage[parfill]{parskip}    		
\usepackage{graphicx}
\graphicspath{{./}}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage[utf8]{inputenc}

@@

All points on $a_i$ within $d_i$ of the $P_{iq}$ are buried, and those bits in $m_{iq}$ are set to zero, with the rest set to 1. This string can be AND'ed with $b_i$ to produce a new $b_i$. This can be done for all spheres that intersect $a_i$ for each sphere in turn. The key insight that cuts down on the $nkl$ calculations is the fact that the $m_{iq}$ can be stored and retrieved later if the same string is ever needed. This is very likely to occur, because the $d_{iq}$ can be made independent of $r_i$ by removing the factor $r_i$. Then $d_{iq}=\sqrt{2(1-\cos\theta_{iq})}$. Now we can calculate $m_{iq}$ for the range of $d_{iq}$ from 0-2. Once we store these masks, we will have covered all possible overlaps between spheres because we have gone through all angles $\theta_{iq}$.
This provides a method of calculating the SASA of the entire protein. We can simply look at every intersection with a given atom and lookup the appropriate mask (based on the angle $\theta_{iq}$. Then we scale it by the radius of the atom we are looking at. Then we can perform our AND operations and find the SASA for a given atom. Iterating over all atoms gives us our answer.

A possible way to improve this method is to increase the number of dots on each atom $l$. Of course, in doing this you trade off time for accuracy. However that tradeoff is a fundamental one in numerically solving almost modeling problem.
\subsection{Richmond's Algorithm and the Gauss-Bonnet Theorem}
In a previous paper written by Richmond (1984), a different strategy is used to find the solvent exposed surface area of the atoms. He still represents the atoms as spheres with definite radius with centers at the atomic coordinates, but instead of using voxels, he made use of the Gauss-Bonnet theorem, a fundamental theorem in differential geometry which connects the topological properties of a manifold (specifically its Euler Characteristic) with its geometric properties (its curvature). A standard statement of the theorem is

@@

\begin{figure}[h!]
\caption{The dual complex ($K$) of the Delaunay Triangulation only includes those power cells that have a non-empty intersection.\cite{Bryant}}
\centerline{\includegraphics[width=3in]{Figure9}}
\end{figure}

@@

\begin{figure}[h!]
\caption{The star of $K$ and link of the star are shown.\cite{Bryant}}
\centerline{\includegraphics[width=3in]{Figure5}}
\end{figure}

@@

where bd $F$ means the boundary of $F$. This equation shows that for the space-filling diagram of the two intersecting spheres, $\sigma_{ij}$ is the fraction of the exposed surface that's on the power cells of both of the spheres.
Using the dual complex, we can further specify the equation for $\sigma_{ij}$ in terms of two intersecting circles, which will become useful later:

@@

The general equation of a sphere ($x^2 + y^2 +z^2 = r^2$) can be used to construct equations for $S_i$ and $S_j$:
\begin{align*}
&x^2 + y^2 +(z-z_i)^2 = \zeta_i^2 \\
&x^2 + y^2 +(z-z_j)^2 = \zeta_j^2 \\
&\text{Now we can combine the equations:} \\
&\zeta_i^2-\zeta_j^2 =(z-z_i)^2-(z-z_j)^2=-2z_i +2z_j +z_i^2 - z_j^2\\
& \implies -2z_i +2z_j +z_i^2 - z_j^2 = (\zeta_i)^2-(\zeta_j)^2\\
\end{align*}
Using this logic, we can find the same equation holds true if we consider the set of balls for each sphere (assuming they have they same radii), as they have the same center as the spheres. So:
\begin{align*}
&-2z_i +2z_j +z_i^2 - z_j^2 = \varrho_i^2-\varrho_j^2 \\
&\text{From the above two expressions, we can state that:} \\
&\zeta_i^2-\zeta_j^2 = \varrho_i^2-\varrho_j^2\\
&\text{In the left term, we can factor out $\zeta_i^2-\zeta_j^2$ to get:} \\
&\zeta_i^2-\zeta_j^2 =(\zeta_i-\zeta_j)(\zeta_i+\zeta_j) = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{So we can state that:} \\
&\varrho_i^2 - \varrho_j^2  = \zeta_{ij}(\zeta_i-\zeta_j) \\
&\text{Now we can find what $\zeta_i$ and $\zeta_j$ really are} \\
&(\varrho_i^2 - \varrho_j^2)/\zeta_{ij} + \zeta_j = \zeta_i \\
&\text{We know that $\zeta_j = \zeta_{ij} - \zeta_i$, so just substitute this expression in:} \\
&\zeta_i = \frac{1}{2}\left[\zeta_{ij}+\frac{\varrho_i^2-\varrho_j^2}{\zeta_{ij}}\right] \\
&\text{A similar statement can be made for $\zeta_j$, following the steps we carried out above:} \\
&\zeta_j = \frac{1}{2}\left[\zeta_{ij}+\frac{\varrho_j^2-\varrho_i^2}{\zeta_{ij}}\right]
\end{align*}

@@

\subsection{Distance Preserving Motion}
In distance preserving motion, one sphere rotates about the other. Imagine three spheres interesting at two points, $p$ and $q$ (Figure 9). This time, $F$ is going to include all the points contained in all three spheres. It is assumed that if the spheres intersect at two points then one sphere contains the other two. In this case, it is $S_j$ that engulfs $S_i$ and $S_k$. The caps, or tops of the spheres, are considered. From a previous section, we know that $B_{ijk}$ is considered the line segment of three intersecting spheres. In this case, $B_{ijk}$ is the line segment containing $p$ and $q$. The caps $C_{ij} = B_i \cup S_j$ and $C_{kj} = B_k \cup S_j$ protrude from the surface of $S_j$. The parameters obtained from this diagram allows us to find the area similar to how we found it for the direction preserving motion. This model says that as you move the two caps apart, keeping their centers in the same position, you will gain an area equal to that of a spherical rectangle. Essentially, the caps are rotating in place. It might be easy to think about how diameter of the caps can represent the height of this “rectangle” while the distance between the caps signifies its width.
\begin{figure}[h!]
\caption{Three intersecting spheres at two points: $S_i$, $S_k$ and $S_j$.\cite{Bryant}}
\centerline{\includegraphics[width=3in]{Figure7}}
\end{figure}

@@

We also plotted the distribution of the SASA values for each calculator, and although they don't look too different in this example, more rigorous testing with a wider range of proteins has shown that \textsc{alphamol} gives far more accurate values than other algorithms, which is why it is incorporated into Rosetta today.
\begin{figure}
\caption{SASA Histograms for the Two Algorithms}
\centerline{\includegraphics[width=3in]{sasa}}
\end{figure}

\section{Conclusion}
It can be seen that SASA is a critical quantity to be able to compute for the purposes of structural biology. On a per-atom basis, it is one metric that can be used to determine if an atom is making a hydrogen bond, and by extension if the structure is energetically favorable. For protein-protein interfaces, large SASA values at the interface usually mean the proteins are more likely to bind because there is a larger binding surface. Metrics like SASA are useful because they don't require your structure to look similar to any existing PDB in order to determine if the structure is likely to be stable. This is especially important in the field of protein engineering, where there might not be any structures to compare your PDB to, and thus having independent metrics is critical to filtering out possible \textit{de novo} protein designs from sequences that won't take on any definite conformation. Thus much time has been spent on applying mathematical and computational ideas to rigorously defining and calculating SASA, and in this paper we have explored a few of the most prominent methods.
